Analyzing DECREASE trials to estimate evidence of data manipulation

### 2017-08-22 submission/manuscript.Rmd
Preprint stage

All manuscript materials are available at https://github.com/chartgerink/2015poldermans and are preserved at Zenodo [@].

@@

To statistically investigate the evidence for data manipulation in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. 
First, we reproduced the findings from the 2014 meta-analysis by Bouri et al. [@bouri2014], which contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also included type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. 
Second, we evaluated the probability that the DECREASE trials (or more extreme effects) arose from the same effect distribution as the non-DECREASE trials, which are assumed to be the true effect of beta-blockers on perioperative mortality in patients undergoing non-cardiac surgery. 
Third, we estimated how many data points would have to be manipulated in order to reproduce the results of the DECREASE trials if the initial, non-manipulated results arose from the effect estimates as obtained from the non-DECREASE trials. Considering the committees investigating the scientific integrity of the DECREASE trials was unable to assess this, we consider it worthwhile to investigate this further.

@@

Based on the effect estimates for the non-DECREASE trials from Step 1, we estimated the probability that the observed effects from the DECREASE studies (or more extreme) occurred naturally. We assumed that the non-DECREASE studies estimated the true effect distribution of perioperative beta-blockade on mortality, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset (9 of 11 results are nonsignificant). Based on this effect distribution, we estimated the veracity of the DECREASE trials separately, which is the estimated probability of the observed data under a given true effect [@peters2015].

@@

Figure `r fignr` indicates that the DECREASE trials are highly unlikely under the estimated effect distribution based on the non-DECREASE trials. More specifically, the results from DECREASE-I (or more extreme) have a probability of 1 in 10 000 ($t(`r df_I`)=`r round(t_I, 3)`,p=`r p_I`$) and the results from DECREASE-IV (or more extreme) have a probability of 1 in 1000 ($t(`r df_IV`)=`r round(t_IV, 3)`,p=`r p_IV`$). This indicates that the DECREASE trial results are unlikely to have come from the same population effect distribution as the non-DECREASE trials. Moreover, observing two of such extremely unlikely results jointly, as in the DECREASE trials, would occur in only 2 out of 10 million sets of two trials (i.e., $`r p_I*p_IV`$) according to this model. Hence, this result indicates that the DECREASE trials are severely different from the non-DECREASE trials.

@@

For DECREASE-IV ($N=`r bouri_data$control_n[11] + bouri_data$beta_n[11]`$), the 95\% confidence interval for the estimated number of manipulated data points is [`r which.min(res_iv_certain - .025) - 1` - `r which.max(res_iv_certain - .975) - 1`] or [`r which.min(res_iv_uncertain - .025) - 1` - `r which.max(res_iv_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The relatively minor difference between the estimates indicates that there is a high degree of confidence that data manipulation did occur based on the difference of the trial results alone. Nonetheless, the range of potentially manipulated data points is still estimated at approximately 1000; this indicates that the summary results are insufficient to provide more than an estimated lower bound. This indicates that it is possible not all data were manipulated (i.e., $N=`r length(res_iv_certain) - 1`$), increasing the importance of well-documented data provenance to discern between genuine and falsified data.

>>>>

All manuscript materials are available at https://github.com/chartgerink/2015poldermans and are preserved at Zenodo [@chris_hartgerink_2017_845355]. CHJH was funded by the Office of Research Integrity during part of this project (ORI; ORIIR160019).

@@

To statistically investigate the evidence for data manipulation in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. 
First, we reproduced the findings from the 2014 meta-analysis by Bouri et al. [@bouri2014], which contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also included type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. 
Second, we evaluated the probability that the DECREASE trials (or more extreme effects) arose from the same effect distribution as the non-DECREASE trials, which are assumed to be the true effect of beta-blockers on perioperative mortality in patients undergoing non-cardiac surgery. 
Third, we estimated how many data points would have to be manipulated in order to reproduce the results of the DECREASE trials if the initial, non-manipulated results arose from the effect estimates as obtained from the non-DECREASE trials. Considering the committees investigating the scientific integrity of the DECREASE trials were unable to assess this, we consider it worthwhile to investigate this further.

@@

Based on the effect estimates for the non-DECREASE trials from Step 1, we estimated the probability that the observed effects from the DECREASE studies (or more extreme) occurred naturally. We assumed that the non-DECREASE studies estimated the true effect distribution of perioperative beta-blockade on mortality, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset (9 of 11 results are nonsignificant). Based on this effect distribution, we estimated the veracity of the DECREASE trials separately, which is the estimated probability of the observed data (or more extreme) under a given true effect [@peters2015].

@@

Results indicate that the DECREASE trials are highly unlikely under the estimated effect distribution from the non-DECREASE trials. More specifically, the results from DECREASE-I (or more extreme) have a probability of 1 in 10 000 ($t(`r df_I`)=`r round(t_I, 3)`,p=`r p_I`$) and the results from DECREASE-IV (or more extreme) have a probability of 1 in 1000 ($t(`r df_IV`)=`r round(t_IV, 3)`,p=`r p_IV`$). This indicates that the DECREASE trial results are unlikely to have come from the same population effect distribution as the non-DECREASE trials. Moreover, observing two of such extremely unlikely results jointly, as in the DECREASE trials, would occur in only 2 out of 10 million sets of two trials (i.e., $`r p_I*p_IV`$) according to this model. Hence, this result indicates that the DECREASE trials are severely different from the non-DECREASE trials.

@@

For DECREASE-IV ($N=`r bouri_data$control_n[11] + bouri_data$beta_n[11]`$), the 95\% confidence interval for the estimated number of manipulated data points is [`r which.min(res_iv_certain - .025) - 1` - `r which.max(res_iv_certain - .975) - 1`] or [`r which.min(res_iv_uncertain - .025) - 1` - `r which.max(res_iv_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The relatively minor difference between the estimates indicates that there is a high degree of confidence that data manipulation did occur based on the difference of the trial results alone. Nonetheless, the range of potentially manipulated data points is still estimated at approximately 1000; this indicates that the summary results are insufficient to provide more than an estimated lower bound. This indicates that it is possible not all data were manipulated (i.e., $N=`r length(res_iv_certain) - 1`$) but at least some were, increasing the importance of well-documented data provenance to discern between genuine and falsified data.
